---
title: "Manakin figure 5 related"
author: "Maggie"
date: "2026-05-11"
output:
  github_document:
    toc: yes
---

This [R Markdown](http://rmarkdown.rstudio.com) Notebook contains codes for reproducing Fig.5A, 5B and Fig.S5A of Balakrishnan et al. 'Genomic and physiological changes in a sexually selected and frugivorous bird radiation'.


```{r, include = FALSE}

library("tidyverse")
library("readxl")
library("gridExtra")
library("grid")
library("RColorBrewer")
library("ggtree")
library("treeio")
library("phytools")
library("viridis")
library("colorspace")

FontSize = 2.5
AxisTxFontSizeSize = 5
AxisTxFontSizeSize_s = 4
AxisTitleFontSizeSize = 5

theme_m =
  theme(
    plot.background = element_blank(),
    plot.title = element_text(hjust = 0.5, size = AxisTitleFontSizeSize),
    panel.grid = element_blank(),
    panel.background = element_blank(),
    panel.border = element_blank(),
    panel.spacing.y = unit(1, "lines"),
    axis.line = element_line(color = "black", size = 0.1),
    axis.title.x = element_text(colour = "black", size = AxisTitleFontSizeSize),
    axis.title.y = element_text(colour = "black", size = AxisTitleFontSizeSize),
    axis.text.x = element_text(colour = "black", size = AxisTxFontSizeSize),
    axis.text.y = element_text(colour = "black", size = AxisTxFontSizeSize),
    axis.ticks = element_line(colour = "black", size = 0.05),
    strip.text = element_text(colour = "black", size = AxisTxFontSizeSize,
                              margin = margin(2, 0, 2, 0, "pt")),
    legend.key = element_blank(),
    legend.text = element_text(colour = "black", size = AxisTxFontSizeSize),
    legend.title = element_text(colour = "black", size = AxisTitleFontSizeSize),
    legend.background = element_blank(),
    legend.position = "none")

set.seed(42)

# setwd('/Users/mbp23/ownCloud/Baldwin/Projects/Manakin/Rscripts/for_github')
```


## Load files
```{r, message=FALSE}
# tree
path = "data/Fig5A_pruned12Jan2021.tre"
bigtree = read.tree(file = path) # 1178

# plumage scores
path = "data/Fig5A_plumage_scores.csv"
sexdim = read_csv(file = path)

# name exchange table (DALE names -> tree tip labels)
path = "data/Fig5A_ManakinNamePlotDale_m.csv"
name_exchange = read_csv(path)

# build sexdim2: adds TipLabel_mod and SexDiff columns
sexdim2 = sexdim %>%
  left_join(name_exchange, by = c("Scientific_name" = "DALE")) %>%
  mutate(tree75sp_mod = ifelse(is.na(tree75sp), Scientific_name, tree75sp),
         TipLabel_mod = gsub(" ", "_", tree75sp_mod),
         SexDiff      = Male_plumage_score - Female_plumage_score)

# combined traits table (diet + morphology)
path = "data/Fig5A_Traits_combined_2020-10-24.csv"
merged4 = read_csv(path)

# join: attach Scientific_name, SexDiff to trait data
diet_sexdim = merged4 %>%
  right_join(sexdim2, by = c("Tip_Label" = "TipLabel"))

# social bonding data
path = "data/Fig5A_preppedSocialBonds.csv"
social = read_csv(path)

diet_sexdim = diet_sexdim %>%
  left_join(social, by = c("Tip_Label" = "species"))

# color scheme (for FigS5A)
path = "data/Fig5A_Diet_ColorSchemes.csv"
colorscheme = read_csv(path)
colorscheme$color_5_202201_alpha = scales::alpha(colorscheme$color_5_202201, 0.7)
```


## Fig5A - Frugivory + SexDiff + social bonding (ape)
```{r, fig.width = 5, fig.height = 12, echo = FALSE}

# convert tip labels from Genus_species to "Genus species"
final_m0 = bigtree
final_m0$tip.label = sapply(final_m0$tip.label, function(s) {
  if (!is.na(s)) paste(unlist(strsplit(s, "_"))[1:2], collapse = " ") else s
})

# attach trait data to tree tips; drop species without frugivory data
trait_data = final_m0 %>%
  as_tibble() %>%
  left_join(diet_sexdim, by = c("label" = "Scientific_name")) %>%
  mutate(TrophicNiche = gsub("Herbivore terrestrial", "Herbivore", TrophicNiche),
         Group = ifelse(grepl("Frugivore", TrophicNiche), "Frugivore", "Non-frugivore")) %>%
  filter(!is.na(Fr))

toremove  = setdiff(final_m0$tip.label, trait_data$label)
tree_trim = final_m0 %>% drop.tip(toremove)

# match frugivory values to tree tip order
ind   = sapply(tree_trim$tip.label, function(s) grep(s, trait_data$label))
trait = trait_data$Fr[ind]
names(trait) = trait_data$label[ind]

# ancestral state reconstruction (frugivory) and color mapping
obj  = contMap(tree_trim, trait, plot = FALSE)
pal  = viridis_pal(option = "inferno", direction = -1)(50)
obj2 = setMap(obj, colors = pal)

# SexDiff heatmap data
temp = trait_data %>%
  filter(!is.na(label)) %>%
  select(label, SexDiff, d)
names(temp)[2:3] = c("SexDiff", "Social")

temp_m = as.matrix(temp$SexDiff)
row.names(temp_m) = temp$label
pal_sex  = sequential_hcl(50, palette = "Blues3", rev = TRUE)
pal_sex2 = colorRampPalette(pal_sex)

# social bonding heatmap data (ordinal: 1 = solitary, 2 = pair, 3 = group)
temp_m2 = as.matrix(temp$Social)
row.names(temp_m2) = temp$label
f = function(n) c("black", "grey40", "grey80")

# plot
plot(obj2, lwd = 0.75, outline = FALSE, fsize = c(0.1, 0.5), ftype = "off", mar = c(10, 1, 1, 8),
     legend = FALSE)
phydataplot(temp_m2, tree_trim, "m", offset = 4, border = NA,
            continuous = FALSE, legend = "none", width = 2, funcol = f)
phydataplot(temp_m,  tree_trim, "m", offset = 6.5, border = NA,
            continuous = TRUE,  legend = "none", width = 2, funcol = pal_sex2)

bar_cex = 0.5  # controls all legend font sizes via par(cex)

par(xpd = NA, cex = bar_cex)  # xpd: draw outside plot; cex: font size for add.color.bar

# frugivory legend
fr_range = range(trait, na.rm = TRUE)
add.color.bar(leg    = 25,
              cols   = obj2$cols,
              title  = "Frugivory",
              lims   = fr_range,
              digits = 1,
              prompt = FALSE,
              x      = par("usr")[1],
              y      = par("usr")[3] + 25,
              subtitle = "")

# plumage SexDiff legend (continuous)
sex_range = range(temp$SexDiff, na.rm = TRUE)
add.color.bar(leg    = 25,
              cols   = pal_sex2(100),
              title  = "Plumage SexDiff",
              lims   = sex_range,
              digits = 1,
              prompt = FALSE,
              x      = par("usr")[1],
              y      = par("usr")[3] - 30,
              subtitle = "")

# social bonding legend (discrete)
legend(x   = par("usr")[1],
       y   = par("usr")[3] - 60,
       legend = c("1", "2", "3"),
       fill   = c("black", "grey40", "grey80"),
       title  = "Social bonding",
       bty = "n", xpd = NA)
```


## FigS5A - Frugivory ancestral state reconstruction + diet breakdown (ggtree)

```{r, fig.width = 6, fig.height = 10, warning = FALSE}

# convert tip labels from Genus_species to "Genus species"
final_m0 = bigtree
final_m0$tip.label = sapply(final_m0$tip.label, function(s) {
  if (!is.na(s)) paste(unlist(strsplit(s, "_"))[1:2], collapse = " ") else s
})

# attach frugivory scores to tree tips; drop species without data
trait_data = final_m0 %>%
  as_tibble() %>%
  left_join(diet_sexdim, by = c("label" = "Scientific_name")) %>%
  mutate(TrophicNiche = gsub("Herbivore terrestrial", "Herbivore", TrophicNiche),
         Group = ifelse(grepl("Frugivore", TrophicNiche), "Frugivore", "Non-frugivore")) %>%
  filter(!is.na(Fr))

toremove  = setdiff(final_m0$tip.label, trait_data$label)
tree_trim = final_m0 %>% drop.tip(toremove)

# match trait values to tree tip order
ind   = sapply(tree_trim$tip.label, function(s) grep(s, trait_data$label))
trait = trait_data$Fr[ind]
names(trait) = trait_data$label[ind]

# ancestral state reconstruction
fit = phytools::fastAnc(tree_trim, trait, vars = TRUE, CI = TRUE)

td = data.frame(node = nodeid(tree_trim, names(trait)), trait = trait)
nd = data.frame(node = names(fit$ace),                  trait = fit$ace)
d  = rbind(td, nd)
d$node = as.numeric(d$node)
tree_trim_m = full_join(tree_trim, d, by = "node")

# color palette
pal = viridis_pal(option = "inferno", direction = -1)(50)

x_limit = max(tree_trim_m@phylo$edge) * 0.021

# tree panel: branches colored by frugivory score
tree_p = ggtree(tree_trim_m, size = 0.25, aes(color = trait),
                continuous = "colour") +
  geom_tiplab(offset = 0.001, size = 0.18, color = "grey50") +
  scale_color_gradientn(colours = pal) +
  xlim_tree(x_limit) +
  theme_tree() +
  theme(legend.position = "bottom")

root_node = ape::Ntip(tree_trim) + 1
tree_p = tree_p %>% ggtree::rotate(root_node)

# diet bar panel: proportions of each dietary category per species
merged5 = merged4 %>%
  mutate(newname = gsub("_", " ", Tip_Label))

target_species = merged5 %>%
  filter(newname %in% tree_trim_m@phylo$tip.label)

traits_m4_long = target_species %>%
  pivot_longer(cols = c(In:Herb),
               names_to  = "TrophicNicheCat",
               values_to = "value2") %>%
  mutate(TrophicNicheCat = gsub("In",   "Invertebrate prey", TrophicNicheCat),
         TrophicNicheCat = gsub("Aq",   "Aquatic prey",      TrophicNicheCat),
         TrophicNicheCat = gsub("Vt",   "Vertebrate prey",   TrophicNicheCat),
         TrophicNicheCat = gsub("Ca",   "Scavenge",          TrophicNicheCat),
         TrophicNicheCat = gsub("Ne",   "Nectar",            TrophicNicheCat),
         TrophicNicheCat = gsub("Fr",   "Fruit",             TrophicNicheCat),
         TrophicNicheCat = gsub("SETe", "Seed",              TrophicNicheCat),
         TrophicNicheCat = gsub("Herb", "Herb",              TrophicNicheCat)) %>%
  select(newname, TrophicNicheCat, value2)

lv_tmp = c("Aquatic prey", "Fruit", "Nectar", "Seed", "Herb",
           "Invertebrate prey", "Scavenge", "Vertebrate prey")
traits_m4_long$TrophicNicheCat = factor(traits_m4_long$TrophicNicheCat, levels = lv_tmp)

col_pal_sm = colorscheme$color_5_202201_alpha[-c(9, 10)]

theme_tmp = theme(
  plot.background   = element_blank(),
  strip.text        = element_blank(),
  axis.text.y       = element_blank(),
  legend.key.size   = unit(8, "points"),
  legend.position   = "right",
  legend.title      = element_text(size = AxisTxFontSizeSize),
  legend.text       = element_text(size = AxisTxFontSizeSize_s),
  legend.background = element_blank())

tree_p2 = tree_p +
  geom_facet(panel = "Diet class break down", data = traits_m4_long, geom = geom_col,
             aes(x = value2, group = TrophicNicheCat, fill = TrophicNicheCat),
             color = "transparent", size = 0.001,
             orientation = "y", width = 1) +
  geom_facet(panel = "Diet class break down", data = traits_m4_long, geom = geom_vline,
             aes(xintercept = 0)) +
  scale_fill_manual(values = col_pal_sm) +
  guides(fill = guide_legend(ncol = 1, byrow = TRUE, reverse = FALSE)) +
  theme_m +
  theme_tmp

tree_p2
```


## Fig5B - Plumage, fruit, and support dot plots

```{r, message=FALSE}
path  = "data/Fig5B_S5D_optima25Feb2022_3_m.csv"
trait = read_csv(path)

trait_sub = trait %>%
  filter(species %in% c("Synallaxis_azarae", "Mionectes_macconnelli",
                        "Pipra_erythrocephala", "Philepitta_castanea"))

trait_sub_l = trait_sub %>%
  pivot_longer(cols = c(plumage.difference,fruit,biparental.care), names_to = "col", values_to = "val")

# trait_sub_l$col = factor(trait_sub_l$col,
                         # levels = c("fruit", "plumage.difference", "biparental.care"))
trait_sub_l$species = factor(trait_sub_l$species, 
                             levels = c("Synallaxis_azarae", "Mionectes_macconnelli",
                                        "Pipra_erythrocephala", "Philepitta_castanea"))
```

```{r, fig.width = 8, fig.height = 2, echo = FALSE}
theme_dot = theme(legend.position = "right",
                  axis.line       = element_blank(),
                  axis.title.x    = element_blank(),
                  axis.title.y    = element_blank())

pr = ggplot(trait_sub_l %>% filter(col == "fruit"),
            aes(x = col, y = species, fill = val)) +
  geom_point(color = "transparent", size = 5, shape = 22, stroke = 1) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "white",
                       limits = c(0, 100)) +
  theme_m + theme_dot

pb = ggplot(trait_sub_l %>% filter(col %in% c("plumage.difference")),
            aes(x = col, y = species, fill = val)) +
  geom_point(color = "transparent", size = 5, shape = 22, stroke = 1) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "white") +
  theme_m + theme_dot + theme(axis.text.y = element_blank())

pg = ggplot(trait_sub_l %>% filter(col == "biparental.care"),
            aes(x = col, y = species, fill = as.factor(val))) +
  geom_point(color = "black", size = 5, shape = 22, stroke = 1) +
  scale_fill_manual(values = c("black", "white")) +
  theme_m + theme_dot + theme(axis.text.y = element_blank())

p = cowplot::plot_grid(pr, pb, pg, nrow = 1, rel_heights = c(3, 1, 1)); p

```

## FigS5D - Plumage, fruit, and support dot plots

```{r}

trait_sub = trait %>%
  filter(species %in% c("Synallaxis_azarae", "Philepitta_castanea", "Calyptomena_whiteheadi",
                        "Pipra_erythrocephala", "Lipaugus_uropygialis",
                        "Mionectes_macconnelli", "Contopus_caribaeus"))

trait_sub_l = trait_sub %>%
  pivot_longer(cols = male.color:Support, names_to = "col", values_to = "val")

trait_sub_l$col = factor(trait_sub_l$col,
                         levels = c("fruit", "female.color", "male.color", "biparental.care", "Support"))
trait_sub_l$species = factor(trait_sub_l$species, 
                             levels = c("Synallaxis_azarae", "Philepitta_castanea", "Calyptomena_whiteheadi",
                        "Pipra_erythrocephala", "Lipaugus_uropygialis",
                        "Mionectes_macconnelli", "Contopus_caribaeus"))
```

```{r, fig.width = 8, fig.height = 4, echo = FALSE}
theme_dot = theme(legend.position = "right",
                  legend.direction = "horizontal",
                  axis.line       = element_blank(),
                  axis.title.x    = element_blank(),
                  axis.title.y    = element_blank())

pr = ggplot(trait_sub_l %>% filter(col == "fruit"),
            aes(y = col, x = species, fill = val)) +
  geom_point(color = "transparent", size = 5, shape = 22, stroke = 1) +
  scale_x_discrete(position = "top") +
  scale_y_discrete(position = "right") +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "white",
                       limits = c(0, 100)) +
  theme_m + theme_dot + 
  theme(axis.text.x.top = element_text(angle = 60, hjust = 0, vjust = 0.5));

pb = ggplot(trait_sub_l %>% filter(col %in% c("male.color", "female.color")),
            aes(y = col, x = species, fill = val)) +
  geom_point(color = "transparent", size = 5, shape = 22, stroke = 1) +
  scale_fill_distiller(palette = "Blues", direction = 1, na.value = "white") +
  scale_y_discrete(position = "right") +
  theme_m + theme_dot + theme(axis.text.x = element_blank())

pg = ggplot(trait_sub_l %>% filter(col == "biparental.care"),
            aes(y = col, x = species, fill = as.factor(val))) +
  geom_point(color = "black", size = 5, shape = 22, stroke = 1) +
  scale_fill_manual(values = c("black", "grey50", "white")) +
  scale_y_discrete(position = "right") +
  theme_m + theme_dot + theme(axis.text.x = element_blank())

# hcl_palettes(plot = TRUE)
trait_sub_l_sup = trait_sub_l %>% filter(col %in% c("Support"))
ps = ggplot(trait_sub_l_sup, aes(y = col, x = species, fill = val)) +
  geom_point(color = "black", size = 5, shape = 22, stroke = 1) +
  scale_fill_gradient(low = 'white', high = '#784421', na.value = 'white', limits = c(0, 100)) +
  scale_y_discrete(position = "right") +
  theme_m + theme_dot + theme(axis.text.x = element_blank())

p = cowplot::plot_grid(pr, pb, pg, ps, ncol = 1, rel_heights = c(3, 2, 1, 1)); p

```